---
title: "Further topics"
author: James McFarland
date: 2019-10-29
output: 
  ioslides_presentation:
    transition: faster
    highlight: haddock
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r, include = FALSE}
library(tidyverse)
library(useful)
library(gapminder)
gapminder <- gapminder %>% 
  mutate(country = as.character(country),
         continent = as.character(continent))
```

## Overview

- Some tips

- Some stats

- 'For loops', and `purrr`


## Practical advice (Rstudio) {.smaller}

- Set up an Rstudio 'project' for each project

  - separate folders for `data`, `code` and `results`
  
  - If need to load data from elsewhere: use the `Import Dataset` tool

- Write analysis like an experiment notebook using .Rmd
  
  - Comments! 

  - Share as html (or pdf)

  - Look for examples you like and use as templates (e.g. `depmap_demo.Rmd`)

  - See the [RMarkdown cheatsheet](https://rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf)


## Practical advice (coding) {.smaller}


- If you're redoing the same thing many times, use a function

- `tidy` data provides a consistent framework that works with `tidyverse` tools

- Making plots: `ggplot` -- Look for examples [online](https://www.r-graph-gallery.com/index.html) and copy the code!

- Google:
  - 'how do I XX in R'
  - 'R error XX'

- Use version control `git`


## Formatting tables {.smaller}

```{r}
library(DT)
datatable(gapminder)
```



## Usesful statistical tests in R

- `wilcox.test()` Wilcoxon rank sum test

- `t.test()` t-test (one or two-sample)

- `cor.test()` test whether correlation is significant (Pearson or Spearman)

- `lm()` linear model

- `aov()`, `anova()` ANOVA (analysis of variance)


## Linear regression {.smaller}

- `dep_var = a * ind_var + b`

- Use funky 'formula' notation: dep_var ~ ind_var

```{r, include = FALSE}
gap_77 <- gapminder %>% filter(year == 1977)
```

```{r}
mod <- lm(lifeExp ~ gdpPercap, data = gap_77) 
```

```{r}
summary(mod) 
```

## Linear regression {.smaller}

- `dep_var = a * ind_var + b`

- Specifying the 'intercept' term explicitly: 

```{r}
mod <- lm(lifeExp ~ 1 + gdpPercap, data = gap_77) 
```

- Without the 'intercept' term:

```{r}
mod <- lm(lifeExp ~ 0 + gdpPercap, data = gap_77) 
```


## Linear regression {.smaller}

- Add more explanatory variables with `+`

```{r}
mod <- lm(lifeExp ~ gdpPercap + pop, data = gap_77) 
summary(mod) 
```


## Linear regression {.smaller}

- Works easily with categorical variables

```{r}
mod <- lm(lifeExp ~ continent, data = gap_77) 
summary(mod)
```

## Linear regression {.smaller}

- Can include 'interaction' terms

```{r}
mod <- lm(lifeExp ~ gdpPercap * continent, data = gap_77) 
summary(mod)
```


## ANOVA {.smaller}

```{r}
mod <- aov(lifeExp ~ continent, data = gap_77)
summary(mod)
TukeyHSD(mod)
```


## For loop {.smaller}

```{r, eval = FALSE}
for (element in vector) {
  ** DO STUFF**
}
```


```{r}
x <- c(2,5,3,9,8,11,6)
count <- 0
for (val in x) {
  count <- count + val
}
print(count)
```

## We've been using them (implicitly)

```{r}
sum(x)
```

```{r}
x + 1
```



## If statement {.smaller}

```{r, eval = FALSE}
if (logical_statement) {
  **DO SOMETHING**
}
```


```{r}
x <- c(2,5,3,9,8,11,6)
count <- 0
for (val in x) {
  if (val %% 2 == 0) {
    count = count+1
  }
}
print(count)
```

## Example loop {.smaller}

Let's compute the median life expectancy for each country in gapminder

```{r}
unique_countries <- unique(gapminder$country) #list to iterate over

med_lifeExp <- vector('numeric', length(unique_countries))
names(med_lifeExp) <- unique_countries
for (cur_country in unique_countries) {
     med_lifeExp[cur_country] <- gapminder %>% 
        filter(country == cur_country) %>% 
        pull(lifeExp) %>% 
        median()
}
head(med_lifeExp)
```


## purrr {.smaller}

Tidyverse tools for doing `for loops`

```{r, eval = FALSE}
result <- map(vector, function)
```

- `map()`: returns a list

- `map_chr()`, `map_int()`, `map_dbl()` and `map_lgl()`: returns a vector of this type (or die trying)

- `map_dfr()` and `map_dfc()` combine results into a dataframe (by row or column)


## Example with purrr

```{r}
get_country_median <- function(cur_country) {
  cur_result <- gapminder %>% 
    filter(country == cur_country) %>% 
    pull(lifeExp) %>% 
    median()
  return(cur_result)
}
med_lifeExp <- map_dbl(unique_countries, get_country_median)
```


## Often a better way {.smaller}
```{r}
gapminder %>% 
  group_by(country) %>% 
  summarise(median = median(lifeExp))
```


## apply

- Apply summary function to each row/column of matrix

```{r, eval = FALSE}
result <- apply(matrix, MARGIN, function, extra_arguments) 
```

- MARGIN = 1 (rows) or 2 (columns)

## apply example {.smaller}

```{r, include = FALSE}
RNAseq_mat <- read_csv(here::here('data', 'CCLE_expression_subset.csv')) %>% 
  column_to_rownames(var = 'X1')
```

```{r}
gene_medians <- apply(RNAseq_mat, 2, median, na.rm=TRUE)
head(gene_medians, 5)
```

## apply with custom functions

Functions must take a vector and return a number

```{r}
get_squared_median <- function(vec) {
  return(median(vec)^2)
}
apply(RNAseq_mat, 2, get_squared_median) %>% head(5)
```

## scale (column-wise normalization) {.smaller}

- 'z-score' normalization is common for matrices

- Subtract off mean of each column

- Divide by its standard deviation

```{r, eval = FALSE}
new_mat <- scale(mat, center, scale)
```

- `center` and `scale` can be either TRUE/FALSE, or a vector to be 'subtracted' or 'scaled' by

## scale example

- mean subtraction

```{r}
norm_RNAseq <- scale(RNAseq_mat, center = TRUE, scale = FALSE)
```

- median subtraction

```{r}
gene_medians <- apply(RNAseq_mat, 2, median, na.rm=TRUE)
norm_RNAseq <- scale(RNAseq_mat, center = gene_medians, scale = FALSE)
```

## Thanks!

More cheat-sheet refs:

[Rstudio2019](https://rstudio.com/wp-content/uploads/2019/01/Cheatsheets_2019.pdf)

[BaseR](https://rstudio.com/wp-content/uploads/2016/10/r-cheat-sheet-3.pdf)
