---
title: "Week 1 Recap"
output: 
  ioslides_presentation:
    transition: faster
    highlight: haddock
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data types

- Vector

- List

- Matrix

- Dataframe

## Functions

```{r, eval = FALSE}
my_function <- function(input1, input2) {
  ...
  return(result)
}
```

```{r, eval = FALSE}
new_stuff <- my_function(stuff1, stuff2)
```


## Loading data

- Text files (comma-separated: `csv`, tab-separated: `tsv`)

- Reading matrices:

  -Use `read.csv`, `read.table`
  
- Reading table:

  -Use `read_csv`, `read_tsv`

- You can always rely on 'Import Dataset' tool in Rstudio!

- Use `here` function to help manage file paths

```{r, eval = FALSE}
df <- read_csv(here::here('data', 'my_file.csv'))
```


## Saving data

- Use the reverse of data loading process

  - `write.csv`, `write_csv`, etc.
  
```{r, eval = FALSE}
write_csv(df, here::here('results', 'my_file.csv'))
```



## Getting data from a vector

- Select data by index:

```{r, eval = FALSE}
idx <- c(2, 5, 2)
vec_sub <- vec[idx]
```

- Select data by logical vector:

```{r, eval = FALSE}
log_idx <- vec > 5
vec_sub <- vec[log_idx]
```

- Select data by name:

```{r, eval = FALSE}
name_idx <- c('first_thing', 'second_thing')
vec_sub <- vec[name_idx]
```


## Getting data from matrices

- Just like subsetting a vector, but subset both rows and columns

```{r, eval = FALSE}
mat_sub <- mat[row_idx, col_idx]
```

- Can leave either row-index, or column-index blank to 'select-all'

```{r, eval = FALSE}
mat_subrows <- mat[row_idx, ]
mat_subcols <- mat[, col_idx]
```


## Getting data from dataframe

- Access one column's worth of data with `$` or 'double-bracket':

```{r, eval = FALSE}
df_sub <- df$my_column
df_sub <- df[['my_column']]
```

- You can access subset of rows treating like a matrix:

```{r, eval = FALSE}
df_rowsub <- df[row_idx, ]
```

- But, **highly** recommend you use `dplyr` to manipulate dataframes



## Dplyr dataframe wrangling

- Key verbs:

  - `Filter`: subset rows
  
  - `Select`: subset columns
  
  - `Arrange`: sort rows
  
  - `Mutate`: add/change column values
  
  - `Summarise`: summarise data across rows
  
  - `Group_by`: group data rows by value to perform grouped-summaries

## Combining tables

- Concatenate rows: `row_bind(df1, df2, ...)`

- Concatenate columns: `col_bind(df1, df2, ...)`

  - Be careful!
  
- Merge tables with a `join`!

  - `inner_join`
  
  - `full_join`
  
  - `left_join`
  
  - `right_join`